---
#title: "Stamped out: In-Person Voting and Covid"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(glmnet)
library(dplyr)
library(readr)
library(lfe)
library(radiant.data)
library(data.table)

fips_df <- read_csv("fips_df.csv")
state_subset <- read_csv("state_subset.csv")
fips_df %>% 
  right_join(state_subset) %>%
  mutate(pre_week_rate = pre_week_diff/pop_2018*100000)-> state_subset

df = state_subset
df = data.table(df)


```
## Stamped Out: In-Person Voting and Covid
I begin by presenting the results of the model and then end with a variety of robustness checks for the models.

### 1. Model Results

#### 1.1 Initial Model & Variable Descriptions
Initial model is presented below:

```{r}

model_turnout <- lm(post_election_rate ~ turnout + pre_election_rate + num_prec_pc + state + med_age +
                         medIncome_2018 + pctPovty_2018 + unplmtRate_2018 + noins_u65_2018 +
                         pctUrban_2010 + pct_vet + avg_householdsize + pct_snap + pctno_hs +
                         pctcollege + popDens + pct_black + pct_hisp + pct_asian +  pct65plus +
                         pct4064 + pct1840 + pctTrump_2016 + pre_week_rate,
                       data = df, weights = pop_2018)

summary(model_turnout)

```

The outcome variable is `post_election_rate`, which is the number of covid cases per 100,000 in the county between November 16th and November 30th. The variable of interest is `turnout`, which is the proportion of a county's residents who cast a vote in-person on Election Day (November 3rd, 2020). 

I control at the county-level for: 

* race [`pct_black`, `pct_hisp`, `pct_asian`],
* education [`pctno_hs`, `pctcollege`], 
* age [`med_age`, `pct65plus`, `pct4064`, `pct1840`]
* socioeconomic status [`medIncome_2018`, `pctPovty_2018`, `unplmtRate_2018`],
* urban [`pctUrban_2010`],
* population density [`popDens`], 
* Trump vote in 2016 [`pctTrump_2016`],
* the number of precincts per capita [`num_prec_pc`],
* health insurance enrollment [`noins_u65_2018`], 
* state fixed effects [stateSTATENAME], 
* and other demographic co-variates like the proportion of veterans [`pct_vet`], 
* the average household size [`avg_householdsize`], 
* and the proportion of county population using SNAP (food stamps) benefits [`pct_snap`].

In addition, I control for a few covid incidence factors: 

* `pre_week_rate`: the relative change in covid rates in the 2 weeks leading up to the election. This accounts for whether covid was increasing or decreasing in a county leading up to election day.
* `pre_election_rate`: the number of covid cases per 100,000 people in a county between October 16 and October 30, before the election

I find significance for `turnout` along with a host of other variables. 
These results suggest that a 1% increase in in-person turnout on Election Day in a county is associated with a 
5.08 case increase per 100,000 people. 

#### 1.2 Simplified Models
I then simplified the model using AIC to obtain a model incorporating the most important predictors. 

In addition, I standardize the co-variates.

```{r, echo = FALSE}

for (var in c('turnout', 'pre_election_rate', 'num_prec_pc', 'pctUrban_2010', 'pct_snap',
              'pct_hisp', 'pctTrump_2016', 'pre_week_rate')) {
  df <- df[!is.na(get(var))]
  df[, (paste0('std_', var)) := ((get(var)) - (weighted.mean(get(var), pop_2018))) /
       (weighted.sd(get(var), pop_2018))]
}

fixed_fx <- felm(post_election_rate ~ std_turnout + std_pre_election_rate + std_num_prec_pc + std_pctUrban_2010 + 
                   std_pct_snap + std_pct_hisp + std_pctTrump_2016 + std_pre_week_rate | state, df, weights = df$pop_2018)
coefficient_plot <- cbind(fixed_fx$coefficients, fixed_fx$se)

coefficient_plot <- cbind(fixed_fx$coefficients, fixed_fx$coefficients - fixed_fx$se, fixed_fx$coefficients + fixed_fx$se)
colnames(coefficient_plot) <- c('mean_estimate', 'low_estimate', 'high_estimate')
rownames(coefficient_plot) <- c('in_person_voting_rate_above_state_avg',
                                'pre_election_covid_cases_per_100k', 'precincts_per_capita',
                                'urban_pct', 'food_stamp_pct', 'hispanic_pct', 'trump_vote_2016', 'pre_election_trend')


```

```{r}

coefficient_plot

```
Above, I present confidence intervals for the change in covid rates for a one-standard deviation
increase in the co-variate. These results suggest that a one-standard deviation increase in the rate of
in-person voting above the state average is associated with 58 additional covid cases
per 100,000 people. 

```{r}

model_turnout = felm(post_election_rate ~ turnout + std_pre_election_rate + std_num_prec_pc + 
                       std_pctUrban_2010 + std_pct_snap + std_pct_hisp + std_pctTrump_2016 + 
                       std_pre_week_rate | state, data = df, weights = df$pop_2018)

summary(model_turnout)
```

These results, with turnout un-standardized, are above. Now, with the model simplified to the most predictive co-variates the effect estimate is 4.95 additional covid cases per 100,000 for a 1% increase in the proportion of a county's residents who voted in-person on Election Day. This is consistent with the full model presented above that found an additional 5.08 covid cases per 100,000.

### 2. Robustness

I then perform a variety of robustness checks to ensure turnout is really a significant predictor.

#### 2.1 State Sensitivity Checks
First, I remove each state in my data, one at a time, to guarantee no single state is
driving these results. For each of the ensuing regressions, I incorporate `pre_election_rate`
using an offset term to more closely match traditional difference in differences. 

```{r, echo = FALSE}

state_list = as.character(unique(df$state))

for(state_ex in state_list){
  df_temp = df %>% filter(state != state_ex)
  
  model = lm(post_election_rate ~ turnout + num_prec_pc + pre_week_rate +  state + 
               med_age + medIncome_2018 + pctPovty_2018 + unplmtRate_2018 + noins_u65_2018 + 
               pctUrban_2010 + pct_vet + avg_householdsize + pct_snap + pctno_hs + pctcollege + 
               popDens + pct_black + pct_hisp + pct_asian +  pct65plus + pct4064 + 
               pct1840 + pctTrump_2016 + offset(pre_election_rate), data = df_temp, weights = pop_2018)
  print(summary(model)[[5]][2,])
}

```
I find significance for the turnout variable in each of the 20 regressions. 
Furthermore, every regression provides an estimate at least as large as 3.75 additional covid cases per 100,000 for each 1% increase in in-person voting on Election Day. Under one specification, the effect is as large as 6.23 additional covid cases per 100,000.

#### 2.2 LASSO Variable Selection
I then use LASSO to select variables with lambda cross-validated.

```{r}
set.seed(07102021)
df = df[complete.cases(df),]
df %>% select(-c(post_covid_cases, pop_2018)) -> X
X = X[,-c(2:10)]
y = df$post_covid_cases

X_test = model.matrix(post_covid_cases ~.-1 - county - FIPS - std_pre_week_rate - state_pre_rate -
                        state_post_rate - state_inperson_rate - std_pre_election_rate - pre_election_rate - 
                        std_turnout - std_num_prec_pc - std_pctUrban_2010 - std_pct_snap -
                        std_pct_hisp - std_pctTrump_2016 - std_pre_week_rate + offset(pre_election_rate),data = df)
X_test = X_test[,-c(21:26,47)]

cvfit = cv.glmnet(X_test,y,alpha = 1, weights = X_test[,3])

lasso_mod = glmnet(X_test, y, alpha = 1, weights = df$pop_2018, lambda = cvfit$lambda.min)
coef(lasso_mod)

```
The results when LASSO selects variables are largely the same. These results suggest that there
are 4 additional covid cases per 100,000 people for a 1% increase in the proportion of people who voted
in person on Election Day.

#### 2.3 County Temperature Checks
I then incorporate October and November temperatures in a county to ensure that these results
are not weather-dependent.

```{r, echo = FALSE, message = FALSE}

temp <- read_csv("temperatures.csv")

df %>% left_join(temp) -> df_temp


model_temp <- lm(post_election_rate ~ turnout + num_prec_pc + state + med_age +
                         medIncome_2018 + pctPovty_2018 + unplmtRate_2018 + noins_u65_2018 +
                         pctUrban_2010 + pct_vet + avg_householdsize + pct_snap + pctno_hs +
                         pctcollege + popDens + pct_black + pct_hisp + pct_asian +  pct65plus +
                         pct4064 + pct1840 + pctTrump_2016 + pre_week_rate + temp_oct + temp_nov + 
                         offset(pre_election_rate), data = df_temp, weights = pop_2018)

summary(model_temp)
```
I retain significance and there are now 3.58 additional covid cases for a 1% increase in the proportion of people
who voted in-person on Election Day. When I instead control for the change in temperature, 
rather than both October and November temperatures, the coefficient for in-person voting on Election Day is 4.85.

#### 2.4 Placebo Check
Lastly, as a placebo check, I perform this same analysis, but on a different time period that does not include the day of the 2020 General Election. I now define the pre-election period as September 16 to September 30 and the "post-election" period as October 16 to October 30 
(rather than October 16 to October 30 and November 16 to November 30, respectively).

With this earlier time period, `turnout` is no longer significant. The direction of the effect is also opposite. 

```{r,echo = FALSE, message = FALSE}

##' Placebo check with old date range to confirm turnout is not significant

county_covid_placebo = read_csv("county_covid_placebo.csv")

df %>% 
  select(-c(pre_covid_cases, post_covid_cases, pre_week_diff)) %>%
  left_join(county_covid_placebo, by = c('FIPS' = 'fips')) %>%
  mutate(pre_election_rate = pre_covid_cases/pop_2018*100000,
         post_election_rate = post_covid_cases/pop_2018*100000,
         pre_week_rate = pre_week_diff/pop_2018*100000) -> df_sept

model_sept = lm(post_election_rate ~ turnout + num_prec_pc + pre_week_rate +  state + 
             med_age + medIncome_2018 + pctPovty_2018 + unplmtRate_2018 + noins_u65_2018 + 
             pctUrban_2010 + pct_vet + avg_householdsize + pct_snap + pctno_hs + pctcollege + 
             popDens + pct_black + pct_hisp + pct_asian +  pct65plus + pct4064 + 
             pct1840 + pctTrump_2016 + offset(pre_election_rate), data = df_sept, weights = pop_2018)

summary(model_sept)

```

